---
title: "Problem Set 1"
author: "Colin Speer"
date: "2024-01-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyr)
```

# Sample data collected from the following paper: 
### Inflammation and Antiviral Immune Response Associated With Severe Progression of COVID-19

*Importing* the **sample** ***data***
```{r data}

sample <- read.table("GSE164805_series_matrix.txt", header = TRUE)

head(sample)

```

Printing the number of rows in the sample
```{r question two}

print(nrow(sample))

```
Printing the number of columns in the sample
```{r question three}

print(ncol(sample))

```
Checking if our sample is stored as a data frame or matrix
```{r question four}

print(str(sample))

```
Checking for NAs in our sample
```{r question five}

if (any(is.na(sample))) {
  print("There are NAs in our data")
} else {
  print("There are no NAs in our data")
}

```
Checking if the **ID_REF** column is a factor
```{r question six}

if (is.factor(sample[["ID_REF"]])) {
  print("The ID_REF column is a factor")
} else {
  print("The ID_REF column is not a factor")
}

```

Setting the **ID_REF** column as a factor
```{r question seven}

sample[["ID_REF"]] <- factor(sample[["ID_REF"]])

if (is.factor(sample[["ID_REF"]])) {
  print("The ID_REF column is a factor")
} else {
  print("The ID_REF column is not a factor")
}

```

Subsetting the data frame using *%in%*
```{r subsetting data frame}

genes <- c("ASHG19AP1B100028737V5", "ASHG19AP1B100017453V5", "ASHG19AP1B100004705V5")
individuals <- c("ID_REF", "GSM5019817", "GSM5019827", "GSM5019828")

subset_sample <- sample[sample$ID_REF %in% genes, individuals]

print(subset_sample)

```

Calculating the mean expression across the three genes for **GSM5019817**
```{r question eight}

print(mean(subset_sample[["GSM5019817"]]))

```

Converting the subset data from **wide** to **long** format
```{r wide to long}

subset_long <- gather(subset_sample, key = "Sample", value = "Expression", -ID_REF, factor_key = TRUE)

print(subset_long)

```

Counting how many observations in our subset data **GSM5019817**, have a gene expression value greater than 4
```{r question nine}

count <- sum(subset_long$Expression > 4, na.rm = TRUE)

print(count)

```